\begin{abstract}
During minimally invasive pulmonary surgery, lung deflation and collapse often make it difficult to precisely remap preoperative CT findings to the intraoperative tissue state. Experienced surgeons may still use locally abnormal stiffness as an important cue for hidden nodule localization, but such tactile judgment is highly operator-dependent and difficult to standardize or quantify. For nodules that are hard to perceive, current practice often relies on invasive localization or costly alternative guidance strategies. The goal of this study was therefore not merely to propose a stronger algorithm, but to build a tactile localization scheme that better matches intraoperative clinical workflow and gradually transforms subjective palpation into measurable and quantitative inference. We constructed a prototype system composed of a flexible tactile array, a front-end acquisition unit, and a workstation, and collected an ex vivo porcine-lung dataset with phantom nodules covering seven sizes, six implantation depths, and three repeated trials. The study then progressed in the same order as the underlying clinical questions emerged: handcrafted features and an XGBoost baseline were first used to verify whether tactile data supported detection, size inversion, and coarse depth discrimination, and to identify task-relevant physical feature families; raw-input hierarchical neural models were then used to directly learn from spatiotemporal tactile tensors; finally, explainability analyses and deployment-oriented integration were used to assess whether the resulting scheme was both interpretable and system-usable. On the independent test set, the raw-input detector achieved an AUC of 0.8383, exceeding the XGBoost baseline of 0.8199. The raw-input size-only model further surpassed XGBoost on size inversion, achieving top-1/top-2/MAE of 0.7177/0.8195/0.1242 versus 0.6701/0.8023/0.1472. For depth, a shared head again collapsed near majority behavior, whereas a size-aware and route-aware raw-input model improved predicted-route balanced accuracy to 0.5240, surpassing the XGBoost depth baseline of 0.5138; a deployment-oriented unified model further improved this value to 0.5337. Explainability analyses indicated that the network encoded part of the physically meaningful structure related to size and depth, although its temporal strategy still relied heavily on the peak neighborhood. Overall, the study supports a cautious conclusion: tactile spatiotemporal sensing can first support reliable intraoperative pulmonary nodule detection, then size inversion, and under explicit hierarchical organization, cautious coarse depth discrimination. The main value of this work lies in establishing a complete problem-solving route from tactile acquisition and repeated experiments to structured validation, raw-input learning, interpretability, and deployment, rather than in reporting a standalone model comparison.
\end{abstract}
